---
title: "Plots of Journey speeds"
author: "Chris Shaw"
date: "19 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overall speed distribution across all routes

```{r overallplot}
hist_norm(journeys$est_speed_mph, 40, "Overall speed across all routes")

```


## Distribution of speed by hour for a busy route

The following show the distribution of speeds for each hour.  The normal distribution with the sample mean and standard deviation is overlaid for comparison

```{r hourplots}
periods <-  c("Morning", "Evening", "Day", "Night")
par(mfrow=c(2,2), oma=c(0,0,2,0))
id<-"SECTIONTL00004"
mtitle=paste0(id, " - ", summary[summary$section_id==id,]$section_description)


route <- journeys[journeys$section_id==id,]

for (p in periods) {
        
        hist_norm(journeys[route$time_period==p,]$est_speed_mph, 40, p)
}

title(mtitle, outer=TRUE)

par(mfrow=c(1,1), oma=c(0,0,2,0))

hist_norm(journeys[journeys$section_id==id,]$est_speed_mph, 40, "Overall distribution of speed for the route")

```



## Plots of overall route speeds

The following show the distribution of speeds for each route.  The normal distribution with the sample mean and standard deviation is overlaid for comparison

```{r sectionplots}
routes <- summary$section_id

for (r in routes) {
        hist_norm(journeys[journeys$section_id==r,]$est_speed_mph, 40, r)
}


```