---
title: "Downloading data from ParlParse"
author: "Chris Shaw"
date: "16 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The raw data is generated by the ParlParse project at:

http://parser.theyworkforyou.com/hansard.html

This is rsynced to a folder on google drive.  Each time this script is run, any new files and any modified files will be updated.  The raw data is **not** synced with github.

Data from year 2000 onward is stored initially.  This could be extended back to 1935 in the future.

```{r filelocations}

sync_dir <- "data.theyworkforyou.com::parldata/scrapedxml/debates/debates2*"
# recommended by parlparse
rsync_opt <- "-az --progress --exclude \'.svn\' --exclude 'tmp/' --relative"
local_dir <- "/cygdrive/c/Users/cs/parl_data"
cygwin_cmd <- "cmd.exe /c c:\\cygwin64\\bin\\run.exe rsync -wait"


system(paste(cygwin_cmd,rsync_opt, sync_dir, local_dir), TRUE)
```

The file names are the only place that reliably contain the date.  The file names need to be converted into date objects

```{r filenamesdates}

debate_data <- "C:\\Users\\cs\\parl_data\\scrapedxml\\debates"
data_files <- list.files(debate_data)


```
