---
title: "Obtaining population data"
author: "Chris Shaw"
date: "28 May 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

# Population data

This file downloads detailed US population data by year state and age group from:

<http://seer.cancer.gov/popdata/download.html>

This is a large dataset which is then processed to obtain relevant summarised datasets:

* population of each state by year

The raw data is obtained from the url if it doesn't already exist:

```{r download, cache=TRUE}

filename<-"us.1969_2014.19ages.adjusted.txt.gz"
data_url <-"http://seer.cancer.gov/popdata/yr1969_2014.19ages/us.1969_2014.19ages.adjusted.txt.gz"

if (!file.exists(filename)){
        download.file(data_url, filename)
}

# Load data and rename data column to something more friendly
#
raw_data<-read.table(filename, stringsAsFactors = F)
names(raw_data)[1]<-"raw"
head(raw_data)

```

The adjusted dataset is used which takes into account the population shifts in certain states following hurricane Katrina.

From time to time the data sets are updated - check and change the data_url above to get the latest data

It can be seen that the raw data is compacted into just one column.  The next section parses and summarises this column.

\newpage 

# Data format

The format of this data is described here:

<http://seer.cancer.gov/popdata/popdic.html>

The key elements needed for this module are:

\begin{center}
    \begin{tabular}{ | l | l | l | p{5cm} |}
    \hline
    Field & Start Col & End Col & Data Type \\ \hline
    Year & 1 & 4 & numeric  \\ \hline
    State & 5 & 6 & char  \\ \hline
    Population & 19 & 26 & numeric  \\ \hline
    \end{tabular}
\end{center}




The following code extracts the relevant columns and caches the results.

```{r parse, cache=TRUE}
raw_data$year<-as.numeric(substr(raw_data$raw, 1,4))
raw_data$state<-as.character(substr(raw_data$raw, 5,6))
raw_data$population<-as.numeric(substr(raw_data$raw, 19,26))

```

# Create a new dataset

In this case we just want population  per state per year:

```{r summarise}

summary <- raw_data %>% group_by(year, state) %>% summarise(population=sum(population))

# write out as a csv
write.table(summary, "states.csv", sep=",", row.names = F)
```

