---
title: "Human Activity Recognition - Analysis of Weight Lifting Exercise"
author: "Chris Shaw"
date: "22 July 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 6, fig.height = 6)
knitr::opts_chunk$set(root.dir="..")
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=" ")
})
setwd("..")
library(knitr)
# initialise analysis
library('ProjectTemplate')
load.project()

```



```{r analysis}
# run the analysis code to generate the objects
knitr::read_chunk('../src/rf-model-caret.R')
 
```

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices enable people to quantify how much of a particular activity they do, but rarely how well they do it. 

This paper analyses a series of  Weight Lifting Exercises carried out by six young health participants who were fitted with a number of such devices.  They performed one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:

* ***Class A*** exactly in accordance with the specification
* ***Class B*** throwing the elbows to the front 
* ***Class C*** lifting the dumbbell only halfway 
* ***Class D*** lowering the dumbbell only halfway 
* ***Class E*** throwing the hips to the front

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.  Measurement data was collected from accelerometers on the belt, forearm, arm, and dumbell.

# Weight Lifting Exercises Dataset

The training dataset contains `r dim(pml.training)[1]` observation samples and `r dim(pml.training)[2]` columns (the last column **classe** is the outcome). 

A

```{r trainingsummary}
# Create a table from the appropriate columns in the training summary data
table_contents<-training_summary[c("start_time", "user_name", "exercise_time", "samples","classe")]

# Add a total row at the bottom
table_contents<-rbind(as.matrix(table_contents), c("Total","", "",
                                                   sum(training_summary$samples), ""))
#Print the table
kable(table_contents,
      row.names = FALSE, col.names = c("Date/time of Exercise", "Participant",
                                       "Time taken (secs)", "Samples Collected", "Class"),
      caption = "Summary of the Weight Lifting training dataset",
      align = c("l", "c", "c","r","c"),
      format.args = list(big.mark=","))
```

# Model

Here are the most 15 important predictors in the model.

```{r plotimpvars, fig.width = 10, fig.height = 6}
varImpPlot(caretrf_mod$finalModel, n.var=15, main="15 most important predictive variables\n in the Random Forest model")
```


